Adaptive group-regularized logistic elastic net regression

### 2019-07-29 docs/biostatistics_main.Rnw
just before resubmit

\externaldocument[sm-]{biostatistics_supp}

@@

	\begin{abstract}
		{In high-dimensional data settings, additional information on the features 
		is often available. Examples of such external information in omics research
		are: (a) $p$-values from a previous study, (b) a summary of prior 
		information, and (c) omics annotation. The inclusion of this information in
		the analysis may enhance classification performance and feature selection,
		but is not straightforward in the standard regression setting. As a solution
		to this problem, we propose a group-regularized (logistic) elastic net
		regression method, where each penalty parameter corresponds to a group of
		features based on the external information. The method, termed 
		\texttt{gren}, makes use of the Bayesian formulation of logistic elastic net
		regression to estimate both the model and penalty parameters in an 
		approximate empirical-variational Bayes framework. Simulations and
		applications in two different molecular marker and cancer type studies show
		that, if the partitioning of the features is informative, classification 
		performance and feature
		selection are indeed enhanced.}
		{Empirical Bayes; High-dimensional data; Prediction; Variational Bayes}
	\end{abstract}

	\section{Introduction}\label{sec:introduction}
	Prediction from high-dimensional data is a challenge common to many fields of
	research. Examples of such high-dimensional prediction problems arise in
	computer vision, stock market prediction and disease diagnosis from omics 
	data. In this paper, we specifically focus on the latter setting. High
	dimensionality of data introduces several issues in the estimation of 
	prediction models, especially: (a) unidentifiable models, (b) highly 
	correlated predictor variables and (c) non-trivial variable selection.

	Attempts to tackle one or more of these issues have lead to the development of
	many prediction methods \cite[]{fan_selective_2010}. Here we will focus on the
	penalization approach. A well-known and popular penalization method is elastic
	net regression \cite[]{zou_regularization_2005}, with its special cases ridge
	regression \cite[]{hoerl_ridge_1970} and lasso regression 
	\cite[]{tibshirani_regression_1996}. The elastic net approach takes issues 
	(a)-(c) into account and has yielded many successful extensions. Recent work 
	on elastic net regression (and its special cases) has focussed on increasing
	prediction accuracy by the inclusion of prior knowledge on the variables
	\cite[]{lee_variable_2017,van_de_wiel_better_2016,tai_incorporating_2007}.

	In omics research, sources of such prior knowledge on the variables are often
	available. The information can, for example, come in the form of (a) results 
	on the same molecular features obtained in a previous study (e.g. $p$-values),
	(b) information from a publicly available database that summarizes the prior
	information on the molecular features involved (e.g., the Cancer Gene Census
	\cite[]{futreal_census_2004}), and (c) omics annotation (e.g. a gene's 
	chromosome location). Although this information can rarely be directly
	included in the statistical analysis, it may still be useful and informative
	for the study at hand.

	Two such studies are introduced in Sections \ref{sec:colorectal} and
	\ref{sec:oral}. The first study is concerned with treatment 
	response prediction for colorectal cancer patients, using sequenced
	microRNAs. In the second study we predict lymph node metastases in oral cancer
	patients, using RNAseq data. In both settings additional information on the 
	features is available in the form of a partitioning. Furthermore, in both
	studies data is sparse, so any extra included information may help to boost
	predictive performance. Inclusion of such extra information in 
	the analysis is not straightforward with standard methods.

	At first sight, such an \textit{a priori} grouping of the features suggests 
	the group lasso
	\cite[]{meier_group_2008, yuan_model_2006}, because
	the group lasso penalises and selects features at the group level. 
	There are two limitations to this approach: the group lasso (a) selects entire
	groups instead of single 
	features and (b) does not penalise adaptively: all groups are penalised 
	equally. Extensions such as the sparse group lasso 
	\cite[]{simon_sparse-group_2013} deal with (a), but do not address (b).

	A natural way to deal with (b) is through differential penalization: that is,
	each group of variables receives its own penalty parameter. An apparent issue
	with differential penalization is the estimation of the penalty 
	parameters. Naive estimation may be done by cross-validation. However, 
	cross-validation requires re-estimation of the model over a grid of penalty
	parameters. The size of this grid increases exponentially with the number of
	penalty parameters. Consequently, it quickly becomes computationally 
	infeasible. We therefore propose an efficient alternative: empirical Bayes
	estimation of the penalty parameters, which corresponds to hyperparameter
	estimation in the Bayesian prior framework. Because of the ubiquity of binary
	outcome data in omics research, we focus on the logistic elastic net.

	In line with the above, our proposed method is (a) objective: the 
	hyper-parameters are estimated from the data and not chosen on subjective 
	grounds, (b) adaptive: prior information is automatically and subjectively 
	weighed with respect to its informativeness, (c) fast compared to full 
	Bayesian analysis or cross-validation, and (d) requires minimal input by the 
	user: only the data and grouping of the features is required. We emphasize 
	that our method is different from recent developments in grouped penalized 
	regression such as \cite{dondelinger_joint_2018}, where instead of the 
	features, the observations are partitioned.

	Recently, \cite{zhang_novel_2019} introduced Variational Bayes Variable
	Selection (VBVS), a method similar to ours in
	spirit. Nevertheless, one clear advantage of our method over VBVS is the
	possibility of adaptive inclusion of external information on the features. 
	Bayesian versions of support vector 
	machines have been used in classification problems as well 
	\cite[]{chakraborty_bayesian_2011}, but these methods lack 
	adaptive inclusion of external information on the features.

	The rest of the paper is structured as follows. We introduce the Bayesian 
	logistic generalised elastic net model in Section
	\ref{sec:model}. In Section \ref{sec:estimation} we derive a variational
	approximation to this model and use this novel approximation in the empirical
	Bayes estimation of multiple, group-specific penalty parameters. In Section
	\ref{sec:simulations} we compare the method in a simulation study. In Sections
	\ref{sec:colorectal} and \ref{sec:oral}, we demonstrate the approach in 
	two data sets. In Section \ref{sec:discussion} 
	we conclude by both discussing some differences and
	parallels between the proposed method, termed \texttt{gren}, and other methods
	in the literature.

	\section{Model}\label{sec:model}
	In logistic regression the outcome variables are assumed to be binary or sums
	of $m_i$ disjoint binary Bernoulli trials ($y_i = \sum_{l=1}^{m_i} k_l, k_l 
	\in \{0,1\}$ for $i=1, \dots, n$). The binomial logistic model relates the
	responses to the $p$-dimensional covariate vectors $\x_i = \begin{bmatrix}
	x_{i1} & \cdots & x_{ip}\end{bmatrix} \tr$ through $y_i \sim \mathcal{B} 
	\( m_i, \expit ( \x_i\tr \bbeta ) \)$, where $\mathcal{B} (m,\upsilon)$ is 
	the binomial distribution with number of trials $m$ and probability 
	$\upsilon$, and $\expit \( \x_i\tr \bbeta \) = \exp(\x\tr_i \bbeta)/[1 + 
	\exp(\x\tr_i \bbeta)]$. Note that if $m_i=1$ for $i=1, \dots, n$, the model
	reduces to a binary logistic regression model. Throughout the rest of the 
	paper we assume that the model matrix $\X = \begin{bmatrix} \x_1 & \cdots &
	\x_n \end{bmatrix}\tr$ is standardized such that $\frac{1}{n}\sum_{i=1}^n
	x_{ij}=0$ and $\frac{1}{n}\sum_{i=1}^n x_{ij}^2=1$ for $j=1, \dots, p$.

	Assume we have a partitioning of the features in $G$ groups, such that each
	feature belongs to one group. Let $\G(g)$ be the feature index set of group 
	$g$ for $g=1, \dots, G$ and let $w_g \in \mathbb{R}_{>0}$ denote a 
	group-specific weight. In a generalised elastic net regression, the penalised 
	likelihood is maximised to yield parameter estimates:
	\begin{align}
	  \hat{\bbeta} & := \underset{\bbeta}{\argmax} \log \L(\y ; \bbeta) - 
	  \frac{\lambda_1}{2} \sum_{g=1}^G \sum_{j \in \G(g)} |w_g \cdot \beta_j| - 
	  \frac{\lambda_2}{2} \sum_{g=1}^G \sum_{j \in \G(g)}\( w_g \cdot \beta_j\)^2 
	  \label{eq:weightedlik}\\
	  & = \underset{\bbeta}{\argmax} \log \L(\y ; \bbeta) -  \frac{\lambda_1}{2} 
	  \sum_{g=1}^G \sqrt{\lambda'_g} \sum_{j \in \G(g)} | \beta_j| - 
	  \frac{\lambda_2}{2} \sum_{g=1}^G \lambda'_g \sum_{j \in \G(g)} \beta_j^2, 
	  \nonumber
	\end{align}
	where $\L(\y ; \bbeta)$ denotes the likelihood function of the 
	observed data $\y = \begin{bmatrix} 
	y_1 & \cdots & y_n \end{bmatrix} \tr$, and 
	$\lambda_1, \lambda_2 \in \mathbb{R}_{>0}$ are the `global' penalty 
	parameters. We wrote $w^2_g = \lambda'_g$ to emphasise that these 
	group-specific weights may be interpreted as penalty multipliers. Note that
	the regular elastic net is recovered by setting $\forall g: \lambda'_g=1$.

  Throughout the following we assume that the geometric mean of the multipliers, 
	weighted by their 
	respective group sizes, is one, such that the average shrinkage of the model
	parameters is determined by $\lambda_1$ and $\lambda_2$. That is,
	we calibrate the $\lambda'_g$ such that $\prod_{g=1}^G 
	(\lambda'_g)^{|\G(g)|}=1$. The multiplier appears in square root form in the
	$L_1$-norm term to ensure that penalisation on the parameter level is the same
	for the $L_1$ and $L_2$-norm terms, as can be seen from the parametrisation in
	(\ref{eq:weightedlik}). The elastic net combines $L_1$ and $L_2$-norm 
	penalisation such that the model parameters are shrunken towards zero. 
	The $L_1$-norm may set some of the estimates exactly to zero, thus 
	automatically selecting features. The $L_2$-norm ensures that correlated 
	features behave similarly, as one would generally require.

	The maximiser of the penalised likelihood in the elastic net, corresponds 
	to the posterior mode of a 
	Bayesian elastic model \cite[]{li_bayesian_2010,zou_regularization_2005}.
	\cite{li_bayesian_2010} show that the elastic net prior 
	(see Supplementary Material (SM) Section \ref{sm-sec:enetprior} for 
	details on this prior) may be written as
	a computationally more convenient scale mixture of normals, with mixing 
	parameter $\btau = \begin{bmatrix} \tau_1 & \cdots & \tau_p \end{bmatrix} 
	\tr$. Using this result, we write the generalised elastic net model in its 
	Bayesian form as:
	\begin{subequations}\label{eq:enmodel2}
		\begin{align}
		\y| \bbeta & \sim \prod_{i=1}^n \mathcal{B} \left( m_i, \expit ( \x\tr_i 
		\bbeta ) \right),  \\
		\bbeta | \btau & \sim \prod_{g=1}^G \prod_{j \in \G(g)} \mathcal{N} \left(0
		,\frac{1}{\lambda' _g \lambda_2} \frac{\tau_j - 1}{\tau_j} \right), \\
		\btau & \sim \prod_{j=1}^p \mathcal{TG} \left( \frac{1}{2},\frac{8 \lambda_2
		}{\lambda_1^2}, \left(1,\infty \right) \right).
		\end{align}
	\end{subequations}
	Here, $\mathcal{TG} \( k,\theta,\( x_l,x_u\) \)$ denotes the truncated gamma 
	distribution with shape $k$, scale $\theta$, and domain $\(x_l,x_u\)$. In this
	Bayesian formulation the penalty parameters $\blambda = \begin{bmatrix} 
	\lambda_1 & \lambda_2 & \lambda'_1 & \cdots & \lambda'_G \end{bmatrix} \tr$ 
	play the role of the hyperparameters in a Bayesian hierarchical model.

	\section{Estimation}\label{sec:estimation}
	\subsection{Empirical Bayes}\label{sec:empiricalbayes}
	If the penalty parameters are known, estimation of the frequentist elastic net
	model parameters or finding the posterior of the generalized elastic net model
	is feasible with small adjustments of the available algorithms 
	\cite[]{friedman_regularization_2010,zou_regularization_2005} or MCMC samplers
	\cite[]{li_bayesian_2010}. Determining these penalty parameters, however, is 
	not straightforward.

	In the frequentist elastic net without group-wise penalisation, two main 
	strategies are used: (i) estimate both $\lambda_1$ and $\lambda_2$ by 
	cross-validation over a two-dimensional grid of values or (ii) re-parametrise 
	the problem in terms of penalty parameters $\alpha= \frac{\lambda_1}{2 
	\lambda_2 + \lambda_1}$ and $\lambda = 2 \lambda_2 + \lambda_1$, fix the 
	proportion of $L_1$-norm penalty $\alpha$ and cross-validate the global 
	penalty parameter $\lambda$. Strategy (i) is advised by 
	\cite{waldron_optimized_2011}, while (ii) is proposed by 
	\cite{friedman_regularization_2010}. In the generalised elastic net setting, 
	strategies (i) and (ii) imply $2 + G$ and $1 + G$ penalty parameters, 
	respectively. $K$-fold cross validation over $D$ values then results in $K 
	\cdot D^{2 + G}$ and $K \cdot D^{1 + G}$ models to estimate. Typically $K$ is 
	set to 5, 10, or to the number of samples $n$, while $D$ is in the order of 
	$100$, so that even for small $G$, the number of models to estimate is 
	prohibitively large.

	In the Bayesian framework, estimation of penalty parameters may be avoided by 
	the addition of a hyperprior to the model hierarchy. The hyperprior takes the 
	uncertainty in the penalty parameters into account by integrating over them. 
	This approach introduces two issues. Firstly, the choice of hyperprior is not 
	straightforward. Many authors suggest a hyperprior from the gamma family of 
	distributions 
	\cite[]{alhamzawi_bayesian_2018,nengjun_yi_bayesian_2013,
	kyung_penalized_2010}, 
	but the precise parametrisation of this gamma prior is not so obvious. A 
	second issue is the loss of correspondence between the  Bayesian and 
	frequentist elastic net. If the ultimate goal is feature selection, this 
	correspondence may be exploited through the automatic feature selection 
	property of the frequentist elastic net. Endowing the penalty parameters with 
	a hyperprior obstructs their point estimation and, consequently, impedes 
	automatic feature selection. Therefore, to circumvent the problem of 
	hyperprior choice and allow for feature selection by the frequentist elastic 
	net, we propose to estimate the penalty parameters by empirical Bayes.

	Many forms of empirical Bayes exist, the most formal one being maximisation of
	the marginal likelihood with respect to the hyperparameters. The resulting 
	hyperparameter estimates are then plugged into the prior. The marginal 
	likelihood is often introduced as a measure of model evidence given the 
	observed data and is computed by integrating the product of likelihood and 
	prior with respect to the model parameters. In the case of the elastic net 
	introduced in (\ref{eq:enmodel2}) this results in the following marginal 
	empirical Bayes posterior for $\bbeta$:
	\begin{align}
	  p_{\hat{\blambda}}(\bbeta | \y) & =
	  \frac{\int_{\btau} \L_{\hat{\blambda}}(\y, 
	  \bbeta, \btau) \, d\btau}{p_{\hat{\blambda}}(\y)} = 
	  \frac{\int_{\btau} \L (\y ; \bbeta) 
	  \pi_{\hat{\blambda}}(\bbeta | \btau) \pi_{\hat{\blambda}}(\btau) \,
	  d\btau}{p_{\hat{\blambda}}(\y)} \nonumber \\
	  \hat{\blambda} & := 
	  \underset{\blambda}{\argmax} p_{\blambda}(\y) = 
	  \underset{\blambda}{\argmax} \int_{\bbeta} \int_{\btau} 
	  \L_{\blambda}(\y, \bbeta, \btau) \, d\bbeta d\btau. \label{eq:mmlestimate}
	\end{align}
	The integrals in (\ref{eq:mmlestimate}) are intractable in the case of the 
	elastic net. In the omics setting the integrals are also high-dimensional, in 
	which case numerical and Monte Carlo approximation methods become tedious and 
	computationally expensive. Moreover, Laplace approximation is known to suffer 
	from low accuracy in many high-dimensional settings 
	\cite[]{shun_laplace_1995}. In \cite{casella_empirical_2001} an EM algorithm 
	is described that estimates the hyperparameters. This EM algorithm iteratively
	maximises the expected joint log likelihood, such that the sequence:
	\begin{equation}\label{eq:ebEM}
	  \blambda^{(k + 1)} = \underset{\blambda}{\argmax} \E_{\bbeta, \btau | \y} \[ 
	  \log \L_{\blambda}(\y, \bbeta, \btau) | \blambda^{(k)} \]
	\end{equation}
	converges to a local maximum of the marginal likelihood. The difficulty herein
	is in the calculation of the expected joint log likelihood. 
	\cite{casella_empirical_2001} suggests to approximate the expectation by its 
	Monte Carlo expectation. Although elegant and simple, this method requires a 
	converged MCMC sample from the posterior for every iteration: a 
	computationally intensive procedure. \cite{roy_selection_2017} introduce
	generalised importance sampling for the Bayesian elastic net, such that we
	just need a limited, pre-specified number of MCMC chains. However, this still
	requires several converged chains, so is not feasible in many high-dimensional
	omics settings. We propose to tackle this problem by approximating the
	expectation in (\ref{eq:ebEM}) using variational Bayes.

	\subsection{Variational Bayes}\label{sec:variationalbayes}
	Variational Bayes is a widely used method to approximate Bayesian posteriors. 
	It has successfully been applied in a wide range of applications, including 
	genetic association studies \cite[]{carbonetto_scalable_2012} and gene network
	reconstruction \cite[]{leday_gene_2017}. In variational Bayes, the posterior 
	is approximated by a tractable form and estimated by optimizing a lower bound 
	on the marginal likelihood of this model (see Section 
	\ref{sm-sec:objectivefunction} of the SM for the 
	lower bound of the proposed model). For an extensive introduction and concise 
	review, see \cite{beal_variational_2003} and \cite{blei_variational_2017}, 
	respectively.

>>>>

\externaldocument[sm-]{supplement}

@@

	\begin{abstract}
		{In high-dimensional data settings, additional information on the features
		is often available. Examples of such external information in omics research
		are: (a) $p$-values from a previous study, and (b) omics annotation. The
		inclusion of this information in
		the analysis may enhance classification performance and feature selection,
		but is not straightforward. We propose a
		group-regularized (logistic) elastic net
		regression method, where each penalty parameter corresponds to a group of
		features based on the external information. The method, termed
		\texttt{gren}, makes use of the Bayesian formulation of logistic elastic net
		regression to estimate both the model and penalty parameters in an
		approximate empirical-variational Bayes framework. Simulations and
		applications to three cancer genomics studies and one Alzheimer metabolomics
		study show that, if the partitioning of the features is informative, 
		classification performance and feature selection are indeed enhanced.}
		{Empirical Bayes; High-dimensional data; Prediction; Variational Bayes}
	\end{abstract}

	\section{Introduction}\label{sec:introduction}
	In cancer genomics studies one is often faced with relatively small sample 
	sizes as compared to the number of features. Data pooling may alleviate the 
	curse of dimensionality, but does not apply to research settings with a unique
	set up, and cannot integrate other sources of information. However, external
	information on the features (e.g. genes) is often ubiquitously available in 
	the public domain. We aim to use this information to improve sparse 
	prediction. The information may come in as feature groups: e.g. the chromosome
	on which a gene is located (24 groups), relation to a CpG island for 
	methylation probes (around 6 groups), or membership of a known gene signature 
	(2 groups). Alternatively, it may be continuous, such as $p$-values from a an
	external study. We introduce a method that allows to systematically use
	multiple sources of such external information to improve high-dimensional
	prediction.

  Our methodology is motivated by four recent, small $n$ clinical omics 
  studies, discussed in detail in Section 
  \ref{sec:simandapp}. The studies concern 
  treatment response prediction for colorectal cancer patients, based on 
  sequenced microRNAs ($n$ = 88); Lymph node metastasis prediction for oral 
  cancer, using RNAseq data ($n$=133); Cervival cancer diagnostics based on 
  microRNAs ($n$=56) and Alzheimer diagnosis based on metabolomics ($n$=87).
  Several
  sources of information on the features were used including: external 
  $p$-values, correlation with DNA markers, the conservation status of microRNAs 
  and node degree in an estimated molecular network.

  As basic prediction model we use the (logistic) elastic net 
  regression \cite[]{zou_regularization_2005}, which combines the 
  desirable properties of its special cases ridge
  \cite[]{hoerl_ridge_1970} and lasso regression 
  \cite[]{tibshirani_regression_1996}: de-correlation and feature selection. 
  It has been demonstrated that the prediction accuracy of penalised regression
  can improve by the inclusion of prior knowledge on the variables. Available
  methods, however, either handle one source of external information only
  \cite[]{lee_variable_2017,tai_incorporating_2007}, or do not aim for sparsity
  \cite[]{van_de_wiel_better_2016}.

  Like others, we assume the external information to be available as feature 
  groups; for continuous information like $p$-values we propose a simple, 
  data-based discretisation (see Section \ref{sec:external}).  At first sight, 
  such \textit{a priori} grouping of the features suggests the group lasso 
  \cite[]{meier_group_2008} 
  or one of its extensions such as the group smoothly clipped absolute 
  deviations (\texttt{grSCAD}) and the group minimax concave penalty 
  (\texttt{grMCP}) 
  \cite[]{huang_selective_2012}. These methods penalise and select 
  features at the group level. This comes with two limitations: the group lasso 
  (a) selects entire groups instead of single features and (b) does not
  penalise adaptively: all groups are penalised
  equally. Extensions such as the sparse group lasso 
  \cite[]{simon_sparse-group_2013} partly deal with (a), but do not address (b).
  Our way to deal with (b) is through differential penalization. That is, each
  group of features receives its own penalty parameter: the group-regularized 
  elastic net (\texttt{gren}). An apparent issue with differential 
  penalization 
  is the estimation of the penalty parameters. Naive estimation may be done by
  cross-validation (CV). However, CV requires re-estimation of the 
  model over a grid, which grows
  exponentially with the number of penalty parameters. Consequently, it quickly
  becomes computationally infeasible. We therefore propose an efficient 
  alternative: empirical-variational Bayes estimation of the penalty parameters,
  which corresponds to hyperparameter estimation in the Bayesian prior 
  framework. Because of the ubiquity of binary outcome data in clinical omics
  research, we focus on the logistic elastic net.

  Recently, \cite{zhang_novel_2019} introduced a method similar to ours as
  it also applies variational Bayes (VB) for feature selection in logistic
  regression. An advantage of our method, however, is the adaptive inclusion of
  external information on the features to aid in prediction and feature 
  selection, as \cite{zhang_novel_2019} do not estimate feature- or 
  group-specific penalty 
  weights. Bayesian versions of support vector machines have been used in
  classification problems as well \cite[]{chakraborty_bayesian_2011},
  but these methods also lack adaptive inclusion of external information on the
  features. In line with the
  above, our proposed method is (a) data-driven: the
  hyperparameters are estimated from the data, (b) adaptive: prior information
  is automatically
  weighted with respect to its informativeness, (c) fast compared to full 
  Bayesian analysis or CV, and (d) easy to use: only the data and
  grouping of the features are required as input.

  The paper is structured as follows. We introduce the model in Section 
  \ref{sec:model}. In Section 
  \ref{sec:external} we shortly discuss possible sources of co-data.
  In Section \ref{sec:estimation} we derive a VB approximation to the
  model introduced in Section \ref{sec:model}
  and use this novel approximation in the empirical Bayes (EB) estimation of 
  multiple, group-specific penalty parameters. In Section \ref{sec:simandapp}
  we compare the method in a simulation study, demonstrate the benefit of the 
  approach for two 
  data sets, and summarise results for two additional data sets. 
  We conclude with a 
  discussion.

	\section{Model}\label{sec:model}
	The outcome variables are assumed to be binary or sums
	of $m_i$ disjoint binary Bernoulli trials ($y_i = \sum_{l=1}^{m_i} k_l, k_l 
	\in \{0,1\}$ for $i=1, \dots, n$). The binomial logistic model relates the
	responses to the $p$-dimensional covariate vectors $\x_i = \begin{bmatrix}
	x_{i1} & \cdots & x_{ip}\end{bmatrix} \tr$ through $y_i \sim \mathcal{B} 
	\( m_i, \expit ( \x_i\tr \bbeta ) \)$, where $\mathcal{B} (m,\upsilon)$ is 
	the binomial distribution with number of trials $m$ and probability 
	$\upsilon$, and $\expit \( \x_i\tr \bbeta \) = \exp(\x\tr_i \bbeta)/[1 + 
	\exp(\x\tr_i \bbeta)]$. Throughout the rest of the 
	paper we assume that the model matrix $\X = \begin{bmatrix} \x_1 & \cdots &
	\x_n \end{bmatrix}\tr$ is standardized such that $\frac{1}{n}\sum_{i=1}^n
	x_{ij}=0$ and $\frac{1}{n}\sum_{i=1}^n x_{ij}^2=1$ for $j=1, \dots, p$.

	Assume we have a partitioning of the features in $G$ groups, such that each
	feature belongs to one group. Let $\G(g)$ be the feature index set of group 
	$g$ for $g=1, \dots, G$ and let $\lambda'_g \in \mathbb{R}_{>0}$ denote a 
	group-specific penalty weight. In a generalised elastic net regression, the
	penalised likelihood is maximised to yield parameter estimates:
	\begin{equation}\label{eq:weightedlik}
	  \hat{\bbeta} := \underset{\bbeta}{\argmax} \log \L(\y ; \bbeta) -  
	  \frac{\lambda_1}{2} 
	  \sum_{g=1}^G \sqrt{\lambda'_g} \sum_{j \in \G(g)} | \beta_j| - 
	  \frac{\lambda_2}{2} \sum_{g=1}^G \lambda'_g \sum_{j \in \G(g)} \beta_j^2, 
	\end{equation}
	where $\L(\y ; \bbeta)$ denotes the likelihood function of the 
	observed data $\y = \begin{bmatrix} 
	y_1 & \cdots & y_n \end{bmatrix} \tr$, and 
	$\lambda_1, \lambda_2 \in \mathbb{R}_{>0}$ are the `global' penalty 
	parameters. From (\ref{eq:weightedlik}) we see that the
	$\lambda'_g$'s may be interpreted as penalty multipliers. Note that
	the regular elastic net is recovered by setting $\forall g: \lambda'_g=1$.

  Throughout the following we assume that the geometric mean of the multipliers,
  weighted by their 
	respective group sizes, is one, such that the average shrinkage of the model
	parameters is determined by $\lambda_1$ and $\lambda_2$. That is,
	we calibrate the $\lambda'_g$ such that $\prod_{g=1}^G 
	(\lambda'_g)^{|\G(g)|}=1$. The multipliers appear in square root form in the
	$L_1$-norm term to ensure that penalisation on the parameter level scales
	with the norm. The $L_1$-norm sets some of the estimates exactly to zero, thus 
	automatically selecting features. The $L_2$-norm ensures collinearity is 
	well-handled.

	The maximiser of the penalised likelihood in the elastic net, corresponds 
	to the posterior mode of a 
	Bayesian elastic model \cite[]{zou_regularization_2005}.
	\cite{li_bayesian_2010} show that the elastic net prior 
	(see Supplementary Material (SM) Section 2 for 
	details) may be written as
	a computationally more convenient scale mixture of normals, with mixing 
	parameter $\btau = \begin{bmatrix} \tau_1 & \cdots & \tau_p \end{bmatrix} 
	\tr$. Using this result, we write the generalised elastic net model in its 
	Bayesian form:
	\begin{subequations}\label{eq:enmodel2}
		\begin{align}
		\y| \bbeta & \sim \prod_{i=1}^n \mathcal{B} \left( m_i, \expit ( \x\tr_i 
		\bbeta ) \right),  \\
		\bbeta | \btau & \sim \prod_{g=1}^G \prod_{j \in \G(g)} \mathcal{N} \left(0
		,\frac{1}{\lambda' _g \lambda_2} \frac{\tau_j - 1}{\tau_j} \right), \\
		\btau & \sim \prod_{j=1}^p \mathcal{TG} \left( \frac{1}{2},\frac{8 \lambda_2
		}{\lambda_1^2}, \left(1,\infty \right) \right).
		\end{align}
	\end{subequations}
	Here, $\mathcal{TG} \( k,\theta,\( x_l,x_u\) \)$ denotes the truncated gamma 
	distribution with shape $k$, scale $\theta$, and domain $\(x_l,x_u\)$. In this
	Bayesian formulation the penalty parameters $\blambda = \begin{bmatrix} 
	\lambda_1 & \lambda_2 & \lambda'_1 & \cdots & \lambda'_G \end{bmatrix} \tr$ 
	play the role of the hyperparameters in a Bayesian hierarchical model.

  \section{External information sources}\label{sec:external}
  We describe possible sources of external information in omics
  studies that may provide the feature groups $\G(g)$. Firstly, there are 
  biologically motivated partitioning of features that are
  easily retrieved from online repositories. These are already 
  discretised and may be included in the analysis as-is. Examples are:
  (i) pathway memberships of genes,
  (ii) classes of metabolites (SM Section 12),
  and (iii) conservation status of microRNAs (SM Section 
  13).
  A second type of external information comes in the form of continuous data.
  Examples are:
  (iv) $p$-values or false discovery rates (FDRs) from a 
  different, related study (Section \ref{sec:colorectal}), and
  (v) quality scores of the features (SM Section 12), 
  and (vii) the node degrees of a network estimated on the feature 
  data (SM Section 12).

  \texttt{gren} requires discretised external information, so in the case of 
  continuous external data, some form of discretisation is required. In some
  cases discretisation comes naturally. For example, with external data 
  type (iv) one might consider the `standard' cut-offs 0.05, 0.01,
  and 0.001. If the choice of cut-offs is not straightforward, we propose a 
  a data-driven heuristic, which renders a relatively finer discretisation grid 
  for data dense than for data sparse domains. Then,
  adaptation to external information is more pronounced in high density areas.
  We propose to fit a piecewise linear spline to the empirical cumulative
  distribution function of the external data and automatically select knot
  locations as cut-offs \cite[]{spiriti_knot_2013}. They also provide a 
  data-driven method to choose the
  number of knots, for a given maximum number of knots. The maximum number
  of knots should be chosen such that each group contains enough features for 
  stable estimation of the penalty weights. As a rule of thumb, we advise
  at least 20 features per group.

  In many practical settings, the external information will be incomplete. 
  We suggest to use a separate
  group of features with missing external information. We prefer this solution
  to setting the penalty multipliers for this group to one, because the absence 
  of external information might be informative from the perspective of 
  prediction.

	\section{Estimation}\label{sec:estimation}
	\subsection{Empirical Bayes}\label{sec:empiricalbayes}
	If the penalty parameters are known, estimation of the elastic net
	model parameters
	is feasible with small adjustments of the available algorithms 
	\cite[]{friedman_regularization_2010,zou_regularization_2005}. Determining 
	these penalty parameters, however, is 
	not straightforward.

	In the frequentist elastic net without group-wise penalisation, two main 
	strategies are used: (i) estimate both $\lambda_1$ and $\lambda_2$ by 
	CV over a two-dimensional grid of values \cite[]{waldron_optimized_2011} or 
	(ii) re-parametrise 
	the problem in terms of penalty parameters $\alpha= \frac{\lambda_1}{2 
	\lambda_2 + \lambda_1}$ and $\lambda = 2 \lambda_2 + \lambda_1$, fix the 
	proportion of $L_1$-norm penalty $\alpha$ and cross-validate the global 
	penalty parameter $\lambda$ \cite[]{friedman_regularization_2010}. In the
	generalised elastic net setting, 
	strategies (i) and (ii) imply $2 + G$ and $1 + G$ penalty parameters, 
	respectively. $K$-fold CV over $D$ values then results in $K 
	\cdot D^{2 + G}$ and $K \cdot D^{1 + G}$ models to estimate. Typically $K$ is 
	set to 5, 10, or to the number of samples $n$, while $D$ is in the order of 
	$100$, so that even for small $G$, the number of models to estimate is 
	very large.

	In the Bayesian framework, estimation of penalty parameters may be avoided by 
	the addition of a hyperprior to the model hierarchy. The hyperprior takes the 
	uncertainty in the penalty parameters into account by integrating over them. 
	This approach introduces two issues. Firstly, the choice of hyperprior is not 
	straightforward. Many authors suggest a hyperprior from the gamma family of 
	distributions 
	\cite[]{alhamzawi_bayesian_2018,
	kyung_penalized_2010}, 
	but the precise parametrisation of this gamma prior is not so obvious. 
	Secondly, the correspondence between the  Bayesian and 
	frequentist elastic net is lost. This 
	correspondence may be exploited through the automatic feature selection 
	property of the frequentist elastic net. Endowing the penalty parameters with 
	a hyperprior obstructs their point estimation and, consequently, impedes 
	automatic feature selection. Therefore, to circumvent the problem of 
	hyperprior choice and allow for feature selection by the frequentist elastic 
	net, we propose to estimate the penalty parameters by EB.

	The most formal form of EB is maximisation of
	the marginal likelihood with respect to the hyperparameters. The resulting 
	hyperparameter estimates are then plugged into the prior. The marginal 
	likelihood is often introduced as a measure of model evidence given the 
	observed data and is computed by integrating the product of likelihood and 
	prior with respect to the model parameters. In the case of the elastic net 
	introduced in (\ref{eq:enmodel2}) EB comes down to finding:
	\begin{equation}\label{eq:mmlestimate}
	  \hat{\blambda} := 
	  \underset{\blambda}{\argmax} p_{\blambda}(\y) = 
	  \underset{\blambda}{\argmax} \int_{\bbeta} \int_{\btau} 
	  \L (\y ; \bbeta) \pi_{\blambda}(\bbeta | \btau) \pi_{\blambda}(\btau) \, 
	  d\bbeta d\btau. 
	\end{equation}
	The integrals in (\ref{eq:mmlestimate}) are intractable in the case of the 
	elastic net. In the omics setting the integrals are also high-dimensional, in 
	which case numerical and Monte Carlo approximation methods become tedious and 
	computationally expensive. Moreover, Laplace approximation is known to suffer 
	from low accuracy in many high-dimensional settings 
	\cite[]{shun_laplace_1995}. In \cite{casella_empirical_2001} an EM algorithm 
	is described that estimates the hyperparameters. This EM algorithm iteratively
	maximises the expected joint log likelihood, such that the sequence:
	\begin{equation}\label{eq:ebEM}
	  \blambda^{(k + 1)} = \underset{\blambda}{\argmax} \E_{\bbeta, \btau | \y} \[ 
	  \log \[ \L (\y ; \bbeta) \pi_{\blambda}(\bbeta | \btau) 
	  \pi_{\blambda}(\btau) \] | \blambda^{(k)} \]
	\end{equation}
	converges to a local maximum of the marginal likelihood. The difficulty herein
	is in the calculation of the expected joint log likelihood. 
	\cite{casella_empirical_2001} suggests to approximate the expectation by its 
	Monte Carlo expectation. Although elegant and simple, this method requires a 
	converged MCMC sample from the posterior for every iteration: a 
	computationally intensive procedure. \cite{roy_selection_2017} introduce
	generalised importance sampling for the Bayesian elastic net, such that we
	just need a limited, pre-specified number of MCMC chains. However, this still
	requires several converged chains, so is not feasible in many high-dimensional
	omics settings. We propose to tackle this problem by approximating the
	expectation in (\ref{eq:ebEM}) using VB.

	\subsection{Variational Bayes}\label{sec:variationalbayes}
	Variational Bayes is a widely used method to approximate Bayesian posteriors. 
	It has successfully been applied in a wide range of applications, including 
	genetic association studies \cite[]{carbonetto_scalable_2012} and gene network
	reconstruction \cite[]{leday_gene_2017}. In VB, the posterior 
	is approximated by a tractable form and estimated by optimizing a lower bound 
	on the marginal likelihood of this model (see SM Section 
	3 for the 
	lower bound of the proposed model). For an extensive introduction and concise 
	review, see \cite{beal_variational_2003} and \cite{blei_variational_2017}.
